---
title: "Multivariate Analysis and Visualization using muvis"
author: "Joseph Oliveira"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(muvis)
library(tidyverse)
data('NHANES')
```

Just following along with the paper, and I get this error.
```{r}
nhanes <- data_preproc(NHANES, is.cat = cat_list, detect.outliers = F,alpha = 0.5)
```

I plucked out the functions I think lead up to this error.

```{r}
nhanes <- data_preproc(NHANES, levels = 15, detect.outliers = F,alpha = 0.5)


Mode <- function(x) {
        ux <- unique(x)
        ux <- ux[which(!is.na(ux))]
        ux[which.max(tabulate(match(x, ux)))]
}


impute.continuous <- function(x) {
        x <- as.numeric(x)
        x[is.na(x)] = mean(x, na.rm = T)
        x
}

impute.factor <- function(x) {
        x <- as.factor(as.character(x))
        x[is.na(x)] = Mode(x)
        x
}

cont.cat.spec <- function(x, is.cat) {
        x <- data.frame(x)
        ls <- c(1:ncol(x))
        t <- sapply(ls, function(i) (!(is.numeric(x[, i]) | is.cat[i])))
        if (sum(t) != 0) {
            ls <- ls[t]
            x <- x[-ls]
            is.cat <- is.cat[-ls]
        }
        x[, is.cat] <- data.frame(sapply(x[, is.cat], as.factor))
        x <- data.frame(x)
        x <- sapply(x, as.numeric)
        x <- data.frame(x)
        if (sum(is.cat) != 0) {
            if (sum(is.cat) > 1) 
                binding <- apply(data.frame(x[, is.cat]), 2, 
                  as.factor)
            else {
                binding <- data.frame(as.factor(t[, which(is.cat == 
                  T)]))
                colnames(binding) <- colnames(x)[which(is.cat == 
                  T)]
            }
            x <- data.frame(x)
            binding <- data.frame(binding)
            x <- cbind(x[, !is.cat], binding)
        }
        x
}

is.cat.function <- function(var) {
        return(!length(unique(var[!is.na(var)])) > levels)
}

```

Apply the function here, and manually work thru to debug.
```{r}
levels = 5
df <- cont.cat.spec(data.frame(NHANES), sapply(data.frame(NHANES), is.cat.function))

alpha = 0.5
df2 <- df[apply(df, 1, function(x) (sum(is.na(x))/length(x)) <= 
        alpha), ]

df2_vec <- c()
nha_vec <- c()
as.data.frame(
  lapply(df2, function(x) {
                            if (is.numeric(x)) 
                            #impute.continuous(x)
                            "num"
                            #append(df2_vec, head(x, 1))
                            else if (is.factor(x)) 
                            "fact"
  }
    )
)

is.factor(df2[,"RIAGENDR"]) 
```

```{r}
length(lapply(NHANES, function(x) if (is.numeric(x) == T) 
    #impute.continuous(x)
    ""
    #append(df2_vec, head(x, 1))
#else if (is.factor(x) == T) 
#    impute.factor(x)
) )  
```

I was able to get down to the root of the error. I can't convert the data into a dataframe, because the `is.factor` in the `lapply` is actually checking a character vector and there the `lapply` returns a `NULL` for that part. 

Checking the difference. The workaround by generating logical vector for is.cat

```{r}

inner_join((df2 %>%
  summarise(across(.fns = function(x) mean(x, na.rm = T))) %>%
  pivot_longer(everything(), names_to = "cols", values_to = "na_ct")),
(NHANES %>%
  summarise(across(.fns = function(x) mean(x, na.rm = T)))%>%
  pivot_longer(everything(), names_to = "cols", values_to = "na_ct")), 
by = "cols") %>%
  mutate(diff = na_ct.x - na_ct.y) %>%
  filter(diff != 0)


data.frame(NHANES)%>%
  select(where(is.character))
df %>%
  select(where(is.character))
df2 %>%
  select(where(is.character))
NHANES %>%
  select(where(is.character))

cat_list <- colnames(NHANES) %in% colnames(df %>%
  select(where(is.character)))


```

```{r}
df2 %>%
  select(RIAGENDR)
```

