---
title: "Multivariate Analysis and Visualization using muvis"
author: "Joseph Oliveira"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(muvis)
library(igraph)

data('NHANES')
```


# Kicking off

When I first ran the preprocessing fucntion below it errored out. I couldn't believe my luck! The paper/package is only 3 years old and it's already incompatible. After digging through the `data_preproc` function and manually reproducing the error, I realized that the fix was to change a logical check on factor type data. The `apply` family of functions strips the factor data type off. What remains is a character vector. So I updated the `is.factor` check to a `is.character` check. Why was this change made to R????

## Preprocessing the data

```{r}
nhanes <- data_preproc(NHANES, levels = 15, detect.outliers = TRUE,alpha = 0.5)
```


```{r}
plot_assoc(nhanes, vars = c("LBXVIE", "LBXTC"))
```

## GGM for Continuous data

The following example constructs a GGM by intersecting the `glasso` and `sin` algorithms. What are these algorithms?

  - ~~Was missing library, `evd`. Installed.~~
  - ~~Was missing dependencies; `evd` and `igraph.` Called explicitly.~~
  - A few functions below are from the `igraph` library:
    - `clusters`, `induced.subgraph`, and `V`

```{r}
nhanes2 <- nhanes
nhanes2$SEQN <- NULL

nhanes2_ggm <- ggm(nhanes2, significance = 0.05, rho = 0.15, community = T, methods = c("glasso", "sin"), plot = F)
grph_clusters <- clusters(nhanes2_ggm$graph)
new_ggm <- induced.subgraph(nhanes2_ggm$graph, V(nhanes2_ggm$graph)[which(grph_clusters$membership == which.max(grph_clusters$csize))])
ggm_vis <- graph_vis(new_ggm, plot = T, filetype = "png", filename = "nhanes_ggm")

```

